---
title: "Get population climate data and plot map"
author: "Alayna Mead"
date: "September 4, 2018"
output: html_document
---


```{r setup}

library(raster)
library(rgdal)
library(ncdf4)
library(ggmap)
library(gridBase)
library(grid)
library(gridExtra)

```


```{r load_and_analyze_flint_data}

# Get data for all variables
# This data is the historical water year data (1951-1980)
# Dataset is from here: http://climate.calcommons.org/node/1129 
# (need an account to sign in)

# put all climate variables in list
clim <- list()

clim[[1]] <- raster('data/flint_climate/aet1951_1980_ave_HST_1534292705.tif')
clim[[2]] <- raster('data/flint_climate/aprpck1951_1980_ave_HST_1534294046.tif')
clim[[3]] <- raster('data/flint_climate/cwd1951_1980_ave_HST_1534292958.tif')
clim[[4]] <- raster('data/flint_climate/pet1951_1980_ave_HST_1534293198.tif')
clim[[5]] <- raster('data/flint_climate/ppt1951_1980_ave_HST_1534293341.tif')
clim[[6]] <- raster('data/flint_climate/rch1951_1980_ave_HST_1534293572.tif')
clim[[7]] <- raster('data/flint_climate/run1951_1980_ave_HST_1534293840.tif')
clim[[8]] <- raster('data/flint_climate/tmn1951_1980_ave_HST_1534294390.tif')
clim[[9]] <- raster('data/flint_climate/tmx1951_1980_ave_HST_1534294244.tif')

names(clim) <- c('aet', 'aprpck', 'cwd', 'pet', 'ppt', 'rch', 'run', 'tmn', 'tmx')

# get information for mother trees

trees <- read.csv('data/phys_data.csv')
trees <- trees[-which(duplicated(trees$mom)),]

coords <- trees[,38:37]

# convert coordinate format to that used in Flint geotiff data

# turn coords into spatial points object
crs <- CRS("+proj=longlat +datum=WGS84") # WGS84 is the default GPS coordinate system
points <- SpatialPoints(coords, proj4string=crs)
# transform from WGS84 to Albers
# info on what parameters to set here: https://proj4.org/operations/projections/aea.html
clim[[1]] # copy parameters  from raster data ('coord. ref.')
points_transform <- spTransform(points, CRS('+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs'))

coordinates(points_transform) # check format
coords2 <- coordinates(points_transform) # rename

# make dataframe for climate data - one row for each mother tree
dat <- as.data.frame(matrix(nrow = nrow(trees), ncol = length(clim)))
colnames(dat) <- names(clim)
rownames(dat) <- rownames(coords2)

# get data for coordinates for all climate variables
for(n in 1:length(clim)){
  dat[,n] <- extract(clim[[n]], coords2)
}

dat$trange <- dat$tmx - dat$tmn
dat$elev <- trees$Elevation.m.

rownames(dat) <- unique(trees$mom)


trees$Site <- factor(trees$Site, labels = c('MC', 'FT', 'FH', 'CV', 'PL', 'RD'), levels = c('MACR', 'TEJO', 'HUNT', 'CENT', 'PLAT', 'REDI'))

cols <- c("#ffe55c", "#ff8b4e", "#e14d66", "#9c2e7f", "#5f0092", "#000262")

par(mfrow = c(2,2))
# look at range in data
# boxplot
for(n in 1:ncol(dat)){
  boxplot(dat[,n] ~ trees$Site, col = cols, main = colnames(dat)[n])
}

# stripchart
# par(mfrow =c(2,2))
#png(file = 'plots/climate_variables_by_site_flint.png', res = 300, height = 6, width = 9, units = 'in')
par(mfrow = c(3,4), mar = c(3,3,3,2))
for(n in 1:ncol(dat)){
  stripchart(dat[,n] ~ trees$Site, vertical = T, method = 'jitter', cex = 1.5, pch = 1, col = cols, main = colnames(dat)[n], ylab = '')
}
#dev.off()

# mostly not too much variation within populations
# PL more variable - has some altitude variation
# exceptions:
# rch, run, tmx variable in PL 
# aet variable in FH, PL, RD


# save in pretty format
mom_flint_climate <- dat
mom_flint_climate$site <- trees$Site
mom_flint_climate$mom <- trees$mom
mom_flint_climate <- mom_flint_climate[,c(11,12,1:10)]

```


```{r load_and_analyze_climate_wna_data}

# get bioclim variables
# these are historical, yearly data for 1961-199
# downloaded from here: https://adaptwest.databasin.org/pages/adaptwest-climatena
# script assumes these are saved in 'data/climate_WNA/'

# convert coordinates
# use projection info here: http://www.cacpd.org.s3-website-us-west-2.amazonaws.com/climate_normals/NA_ReadMe.txt
# following: https://stackoverflow.com/questions/30287065/convert-lambert-conformal-conic-projection-to-wgs84-in-r

crs <- CRS("+proj=longlat +datum=WGS84")
p <- SpatialPoints(coords, proj4string=crs)
g <- spTransform(p, CRS("+proj=lcc +lat_1=49 +lat_2=77 +lat_0=0 +lon_0=-95 +datum=WGS84"))

coordinates(g)

# import all variables
vars <- c('AHM', 'bFFP', 'CMD', 'DD_0', 'DD_18', 'DD18', 'DD5', 'eFFP', 'EMT', 'Eref', 'EXT', 'FFP', 'MAP', 'MAR', 'MAT', 'MCMT', 'MSP', 'MWMT', 'NFFD', 'PAS', 'PPT_sm', 'PPT_wt', 'RH', 'SHM', 'Tave_sm', 'Tave_wt', 'TD')

# make list of variables
clim_nc <- list()
for(n in 1:length(vars)){
  clim_nc[[n]] <- raster(paste('data/climate_WNA/', vars[n], '.nc', sep = ''))
}
names(clim_nc) <- vars

# extract data from raster for coordinates
dat_nc <- as.data.frame(matrix(ncol = length(vars), nrow = nrow(coords)))
colnames(dat_nc) <- vars
rownames(dat_nc) <- trees$mom
for(n in 1:length(vars)){
  dat_nc[,n] <- extract(clim_nc[[n]], coordinates(g))
  print(c('done with var', n), quote = F)
}

# look at range in data
# boxplots
par(mfrow = c(2,2))
for(n in 1:ncol(dat_nc)){
  boxplot(dat_nc[,n] ~ trees$Site, col = cols, main = colnames(dat_nc)[n])
}

# stripchart
#par(mfrow =c(2,2))

#png(file = 'plots/climate_variables_by_site_climate_wna.png', res = 300, height = 9, width = 13, units = 'in')
par(mfrow = c(5,6), mar = c(3,3,3,2))
for(n in 1:ncol(dat_nc)){
  stripchart(dat_nc[,n] ~ trees$Site, vertical = T, method = 'jitter', cex = 1.5, pch = 1, col = cols, main = colnames(dat_nc)[n])
}
#dev.off()

# again not too much variation
# PL somewhat variable in bFFP, all DD vars, MAT, MWMT, Tave_sm

```

```{r save_data}

# combine flint and climate wna datasets
climate <- cbind(dat, dat_nc)
climate$mom <- rownames(climate)
climate <- climate[c(39,1:38)]

# save
write.csv(climate, file = 'data/climate_data_by_mom_tree_flint_and_climatewna.csv', quote = F, row.names = F)


```


```{r map}
 
# get map
map_bw <- get_stamenmap(c(top = 42.3, right = -113.5, bottom = 32, left = -125), maptype = 'toner-background', zoom = 6)
ggmap(map_bw)

# set up for plotting site points
sites <- coords[c(4,11,16,21,27,33),] # get one coordinate for each site

cols <- c("#ffe55c", "#ff8b4e", "#e14d66", "#9c2e7f", "#5f0092", "#000262")
cols2 <- cols[c(4,3,2,1,5,6)] # reorder colors

# plot map
m <- ggmap(map_bw) + labs(x = '', y ='')
m <- m + scale_x_continuous(limits = c(-124, -114))
m <- m + geom_point(aes(x = Longitude, y = Latitude), data = sites, alpha = 1, show.legend = F, size = 5,  fill = cols2, color = 'grey10', shape = 21, stroke = 1.3 )
m <- m + geom_text(data = sites, aes(x = Longitude, y = Latitude, label = c('Centerville','Fort Hunter Liggett', 'Fort Tejon', 'Malibu Creek','Platina','Redding')), color = 'grey10', size = 7, hjust=-.1, vjust=.5, show.legend = F)
plot(m)

ggsave(file = 'plots/map_with_sample_locations.png', device = 'png', width = 8, height = 8, units = 'in', dpi = 300)

t.ax <- 10 # axis text size
t.axt <- 12 # axis title text size
t.m <-  14 # main title text size

# barplots of climate data
p1 <- ggplot(as.data.frame(aggregate(climate$MAT, by = list(trees$Site), mean)))
p1 <- p1 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p1 <- p1 + coord_cartesian(ylim=c(14,18))
p1 <- p1 + labs(title = 'Mean Annual\nTemperature')
p1 <- p1 + ylab(expression(~degree~ C))
p1 <- p1 + xlab('')
p1 <- p1 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p1)

p2 <- ggplot(as.data.frame(aggregate(climate$tmx, by = list(trees$Site), mean)))
p2 <- p2 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p2 <- p2 + coord_cartesian(ylim=c(20,26))
p2 <- p2 + labs(title = 'Maximum\nTemperature')
p2 <- p2 + ylab(expression(~degree~ C))
p2 <- p2 + xlab('')
p2 <- p2 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p2)

p3 <- ggplot(as.data.frame(aggregate(climate$MSP, by = list(trees$Site), mean)))
p3 <- p3 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p3 <- p3 + coord_cartesian(ylim=c(10,100))
p3 <- p3 + labs(title = 'May-September\nPrecipitation')
p3 <- p3 + ylab('mm')
p3 <- p3 + xlab('')
p3 <- p3 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p3)

p4 <- ggplot(as.data.frame(aggregate(climate$cwd, by = list(trees$Site), mean)))
p4 <- p4 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p4 <- p4 + coord_cartesian(ylim=c(600,1150))
p4 <- p4 + labs(title = 'Climatic Water\nDeficit')
p4 <- p4 + ylab('mm')
p4 <- p4 + xlab('')
p4 <- p4 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p4)

p5 <- ggplot(as.data.frame(aggregate(climate$AHM, by = list(trees$Site), mean)))
p5 <- p5 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p5 <- p5 + coord_cartesian(ylim=c(25,95))
p5 <- p5 + labs(title = 'Annual Heat\nMoisture Index')
p5 <- p5 + ylab('')
p5 <- p5 + xlab('')
p5 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p5)

p6 <- ggplot(as.data.frame(aggregate(climate$elev, by = list(trees$Site), mean)))
p6 <- p6 + geom_point(aes(x=Group.1, y=x), color = cols, size = 3)
p6 <- p6 + coord_cartesian(ylim=c(100,1000))
p6 <- p6 + labs(title = 'Elevation')
p6 <- p6 + ylab(expression('M'))
p6 <- p6 + xlab('')
p6 <- p6 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text(size = t.m, hjust = 0.5), axis.text = element_text(size = t.ax, color = 'black'), axis.title = element_text(size = t.axt, color = 'black'))
print(p6)

########################################################
# plot map and plots together

# set up matrix arrangement
mat <- matrix(c(1,1,1,2,5,
                1,1,1,3,6,
                1,1,1,4,7),
              ncol = 5, byrow = T)


# just plot
grid.arrange(m,p1,p3,p5,p2,p4,p6, layout_matrix = mat)

# save to grob, then to file
all <- arrangeGrob(m,p1,p3,p5,p2,p4,p6, layout_matrix = mat, vp=viewport(width=0.95, height=1))

ggsave(file = 'plots/map_and_climate_variables.pdf', plot = all, device = 'pdf', width = 11, height = 7, units = 'in')

# note: edited water color in gimp

```

Climate variable abbreviations

################################################################
Climate WNA:

1) Annual variables:

Directly calculated annual variables:

MAT              mean annual temperature (°C),

MWMT           mean warmest month temperature (°C),

MCMT            mean coldest month temperature (°C),

TD                   temperature difference between MWMT and MCMT, or continentality (°C),

MAP               mean annual precipitation (mm),

MSP                May to September precipitation (mm),

AHM  annual heat-moisture index (MAT+10)/(MAP/1000))

SHM               summer heat-moisture index ((MWMT)/(MSP/1000))

           

Derived annual variables:

DD<0              degree-days below 0°C, chilling degree-days

DD>5              degree-days above 5°C, growing degree-days

DD<18            degree-days below 18°C, heating degree-days

DD>18            degree-days above 18°C, cooling degree-days

NFFD              the number of frost-free days

FFP                 frost-free period

bFFP               the day of the year on which FFP begins

eFFP                the day of the year on which FFP ends

PAS                 precipitation as snow (mm) between August in previous year and July in current year

EMT                extreme minimum temperature over 30 years

EXT                extreme maximum temperature over 30 years

Eref                 Hargreaves reference evaporation (mm)

CMD               Hargreaves climatic moisture deficit (mm)

MAR               mean annual solar radiation (MJ m‐2 d‐1)

RH                  mean annual relative humidity (%)

 
2) Seasonal variables:

Seasons:

Winter (_wt): Dec. (prev. yr) - Feb for annual, Jan, Feb, Dec for normals

Spring (_sp): March, April and May

Summer (_sm): June, July and August

Autumn (_at): September, October and November 

###############################################################
Flint dataset:

Actual evapotranspiration - aet
April 1 snowpack - aprpck
Climatic water deficit - cwd
Maximum monthly temperature - tmx
Minimum monthly temperature - tmn
Potential evapotranspiration - pet
Precipitation - ppt
Recharge - rch
Runoff - run
